---
title: "sdmTMB Sardine eggs Bay of biscay "
format:
  html:
    toc: true
    toc-depth: 4
    toc-float: true
    code-fold: true         
    code-summary: "Hide/Show "
    code-tools:
      toggle: true         
---

### Load libraries and data

```{r,include=F}
rm(list=ls())

# Load packages
library(tidyverse)
library(dplyr)
library(ggplot2)
library(mapdata)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(sp)
library(sdmTMB)

# Map
world <- ne_countries(scale = "medium", returnclass = "sf")
```

```{r,warning=F}
variable='Sp10m3'; species='Sardine'; ccolor='blue';maxvalue=200
listyears=c(2000:2019,2021:2024)
file.data='DataEggs/Cufes_spring_ifremer.txt'
data.eggs=read.table(file=file.data,sep="\t",header=T)
```

#### A quick map of your row data

```{r,warning=F}
data.eggs_sf <- st_as_sf(data.eggs, coords = c("Long", "Lat"), crs = "WGS84")
plot_eggs <- ggplot(data = data.eggs_sf) + 
  geom_sf(aes(color = Sp10m3), alpha = 0.8, size = 0.8) +
  facet_wrap(~ Year) +
  scale_color_viridis_c(trans = "log", na.value = "grey80",   name = expression("Log(Eggs/10m³)\n grays are zero")) +
ggtitle("Total Sardine eggs")
plot_eggs
```


Those are the row data of eggs from PELGAS survey for years 2000-2024 (1 year -2020- gap). Eggs are collected using CUFES at 5m depth, every 3nmi (5.556Km) at transects with inter-transect distance of 12nmi (22.22Km).

Sp10m3 & Ee10m3: total eggs **density** (no eggs/ 10m ^3^ ) for sardine and anchovy respectively.

Sp1,2,3 & Ee1,2,3: the **number** of eggs or sardine and anchovy respectively for specific stages.

Group1: no embryo, egg stages (1-3)

Group2: early embryo, egg stages (4-6)

Group3: late embryo, egg stages (7-11).

Sonde(depth), Ventvrai, Dirventvrm, Tempsbe Salinite, chlorophyl,

*We have to think:*

-   *The vertical distribution of eggs*

-   *Eggs cohorts (Total eggs, group1, aging and Po) [@petitgas2009]* at least remove the effect of temperature by standardizing with stage duration time (Renger et al 1985) 

-   * We removed data at the north of 48°N*

```{r,warning=F}
data.eggs<- data.eggs %>% filter(Lat < 48 & is.na(Sp10m3)==F) %>%
  mutate(Year=as.factor(Year), time=as.integer(Year))
data.eggs <- add_utm_columns(data.eggs, c("Long", "Lat"), ll_crs = 4326, units = "km")   #Project coordinates to Km
```

#### A quick map of your final data

```{r,warning=F}
data.eggs_sf <- st_as_sf(data.eggs, coords = c("Long", "Lat"), crs = "WGS84")
plot_eggs <- ggplot(data = data.eggs_sf) + 
  geom_sf(aes(color = Sp10m3), alpha = 0.8, size = 0.8) +
  facet_wrap(~ Year) +
  scale_color_viridis_c(trans = "log", na.value = "grey80",   name = expression("Log(Eggs/10m³)\n grays are zero")) +
ggtitle("Total Sardine eggs")
plot_eggs
```

### Prepare grid for prediction

We are going to need another grid in which we will predict the eggs abundance using the final model.\
The prediction grid has to be replicated for the years included in the study; it also needs to have all the covariates that were used in the model. To have these variables in the spatial resolution of the prediction grid instead of an classic interpolation here we use same approach as[@lindmark2023] .

Based on than approach you built a simple spatio-temporal model for each covariate (ex. m_temperature: μ~s,t~= β~t~+ ω~s~ +ε~s,t~) of your final model and you use that model (m_temperature) to predict the variable (here temperature) in the resolution of the prediction grid. You repeat that process for all your covariates.

So, finaly you will you have the final prediction grid, which has information of all the fixed covariates included in your final model in the spatial resolution of the prediction grid.

```{r,warning=F}
#Insert the grid
projection_grid <- readRDS("./DataEggs/user_region.rds")
projection_grid_UTM <- add_utm_columns(projection_grid, c("Lon", "Lat"), ll_crs = 4326, units = "km")
plot(projection_grid_UTM$X,projection_grid_UTM$Y)
```

```{r,warning=F}
#Replicate the grid across year
grid_yrs <- replicate_df(projection_grid_UTM, c("Year"), as.factor(unique(data.eggs$Year))) %>%
  mutate(time=as.integer(Year))

```

### Create the mesh

In these kinds of models, the spatial (ω~s~) and the spatiotemporal (ε~s,t~) variations (and potentially any spatial varying coefficient) are modeled as Gaussian Random Fields (GRF), where the random effects describing the spatial patterns are assumed to be drawn from a multivariate normal distribution MVNormal (0,Σ). Estimating this covariance matrix (Σ) is computational demanding, so the SPDE approach is used to approximate the Gaussian random fields.

For the SPDE approximation approach you need another grid called the **mesh** which is a triangulated mesh with Knots. These Knots are used to approximate the spatial variability in your data and because the number of Knots \<\< number of data points the computation of the covariance is easier. The Matern correlation is used as a covariance function to account for spatial autocorrelation.

*Maybe more discussion in how you contract the mesh (ex. who many Knots you need)*

So, so far we have three grids with different spatial resolution :

-   The data grid (your sampling points, those data are used to estimate fixed parameres)

-   The mesh (for the Gaussian Random Fields approximation)

-   The prediction grid ( for predictions using your final model).

    So the mesh is:

```{r warning=FALSE}
mesh <- make_mesh(data.eggs, xy_cols = c("X", "Y"), cutoff = 30)  # cutoff: minimum distance between knots before a new mesh vertex is added- units as in your dataset
plot(mesh)                                                     
```

## MODEL 1 (m~1~) : Spatial model (non spatio-temporal term)

The first model (m~1~) follows that formula:

μ~s,t~= exp(β~t~+ ω~s~),

μ~s,t~= log(y~s,t~ ),

y~s,t~ \~ Tweedie (μ~s,t~ ,p, φ) , ω\~ MVNormal (0,Σ~ω~)

where the modeled egg density (μ~s,t~) follows the Tweedie distribution with a log link function, as egg density is positive, continuous and contains zeros.

y~s,t~ represents eggs density (no of eggs/10m^2^) at space (s) and time(t), μ is the mean density at time and space and p, φ represent power and dispersion parameters of the Tweedie distribution, respectively.

β~t~ are the fixed effects for factor(Year), they represent the average temporal variation

ω~s~ is the spatial random effect, drawn from a GRF with covarianvce matrix Σ~ω~ using Maretn correlation function. It represents the average spatial pattern (nich effect) and it also accounts for spatial variation that is not accounted for by our fixed effects. In other words, these deviations represent consistent biotic and abiotic factors that are affecting biomass density but are not accounted for in the model.

### Fit the model

```{r,warning=F}
fit_m1 <- sdmTMB(
  Sp10m3 ~ 0 + Year,
  data = data.eggs,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "on"
)

```

### Results of the fit

```{r,warning=F}
fit_m1  
```

#### Confidence intervals of fixed effects

```{r,warning=F}
fix_effects<- tidy(fit_m1, conf.int = TRUE)
fix_effects
```

#### Confidence intervals of random effect and variance parameters

```{r,warning=F}
random_effects<- tidy(fit_m1, effects = "ran_pars", conf.int = TRUE)
random_effects
```

Other than the effect of the spatial random field you have: :

-   Matérn range: the distance at witch you have no correlation - similar to the variogram range

-   phi (φ) & tweedie_p (p): the dispersion parameter and the power of the Tweedie distribution, respectively.

##### Plot the all the effects

```{r,warning=F}
ggplot() +
  geom_point(data = fix_effects, aes(x = estimate, y = term, col = term)) + 
  geom_errorbar(data = fix_effects, aes(y = term, xmin = conf.low, xmax = conf.high, color = term), width = 0.2) +  
  geom_point(data = random_effects %>% filter(term == "sigma_O"), aes(x = estimate, y = term)) +  
  geom_errorbar(data = random_effects %>% filter(term == "sigma_O"), aes(y = term, xmin = conf.low, xmax = conf.high), width = 0.2) +  
  labs( title = expression("Mean and 95% confidence intervals of the fixed & random effects," ~ m[1]),       
        subtitle = expression(mu[s~t] ~ "=" ~ exp(beta[t] + omega[s]))) +
  theme_bw() +
  theme(legend.position = "none")
```

You can see that the effect of spatial term is the most "important" one. Space is the more influential parameter in egg distribution in this model.

### Diagnostics

What do you check?

Residuals?

```{r,warning=F}
sanity(fit_m1)
```

### Prediction

You now perform a prediction in you prediction grid using the fitted model m~1~. When you visualize your prediction in a map you can visualize your final prediction your model (eg. the estimated number of eggs) or you can visualize separately each part of the decomposition (the effect of the fixed covariates and the effect of the random fields (ω~s~, ε~s,t~) and see what is going on.

```{r,warning=F}
#Predict in the prediction grid that has been created above
p_m1 <- predict(fit_m1, newdata = grid_yrs)
```

#### Full (fixed and random effects) model m~1~ prediction

```{r,warning=F}
plot_m1 <- ggplot(data=p_m1, aes(X, Y, fill = exp(est))) + 
  geom_raster() + facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (fixed effects +  random effects) based on full m1",
subtitle = expression("The estimated density of eggs " * log(mu[st]) * ": " * mu[st] == exp(beta[t] + omega[s])))

plot_m1
```

#### Fixed effect in the of prediction model m~1~

```{r,warning=F}
plot_m1 <- ggplot(data=p_m1, aes(X, Y, fill = exp(est_non_rf))) + 
  geom_raster() + facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (fixed effects only) based on full m1",
          subtitle = expression("The effect of time in the estimated variable " * (beta[t]) * ": " * mu[st] == exp(beta[t] + omega[s])))
plot_m1
```

#### Random effect in the prediction of model m~1~

Here spatial random effect only

```{r,warning=F}
plot_m1 <- ggplot(data=p_m1, aes(X, Y, fill = exp(omega_s))) + 
  geom_raster() +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (random effects only) based on full m1",
          subtitle = expression("The effect of space in the estimated variable " * (omega[s]) * ": " * mu[st] == exp(beta[t] + omega[s])))
plot_m1
```

For the ω~s~ we have only one map because the spatial effect is stable across time. This is the mean spatial effect estimated from your data and it also represents the mean spatial distribution. If you want to see if or how the spatial effect changes in time then you need a spatiotemporal random effect (see m~2~).

This spatial random effect represents variance in space (but consistence through time) coming from the data that are not accounted by the fixed effects. In this model compared to m~1~ the spatial random effect is less influential because a part of the variability is now explained by the spatiotemporal effect term.

## MODEL 2 (m~2~) : Spatiotemporal model

The 2^nd^ model (m~2~) follows that formula:

μ~s,t~= exp(β~t~+ ω~s~ + ε~s,t~),

μ~s,t~= log(y~s,t~ ),

y~s,t~ \~ Tweedie (μ~s,t~ ,p, φ) , ω\~ MVNormal (0,Σ~ω~) , ε~t~ \~ MVNormal (0,Σ~t~)

where the modeled egg density (μ~s,t~) follows the Tweedie distribution with a log link function, as egg density is positive, continuous and contains zeros.

y~s,t~ represents eggs density (no of eggs/10m^2^) at space (s) and time(t), μ is the mean density at time and space and p, φ represent power and dispersion parameters of the Tweedie distribution, respectively.

β~t~ are the fixed effects for factor(Year), they represent the average temporal variation

ω~s~ is the spatial random effect, drawn from a GRF with covarianvce matrix Σ~ω~ using Maretn correlation function. It represents the average spatial pattern (nich effect) and it also accounts for spatial variation that is not accounted for by our fixed effects. In other words, these deviations represent consistent biotic and abiotic factors that are affecting biomass density but are not accounted for in the model.

ε~s,t~ represents the spatiotemporal random effect, drawn from a GRF with covarianvce matrix Σ~ε~. They represents how the average spatial pattern change through year so they also for spatial variation that is not accounted for by our fixed effects and the spatial random effects. In other words, these deviations represent consistent biotic and abiotic factors that are affecting our study variable but are not accounted for in the model.

In this model Spatiotemporal random fields are considered independent and identically distributed each single year (spatiotemporal = "IID") and Maretn correlation function to accounts for spatial autocorrelation.

*Are there other options for spatiotemporal fields:*

-   *"AR1": first-order autoregressive (each random filed is correlated to the one of the previous year )*

-   *"RW": random walk (each random filed starts fro where it was last year plus a completely independent deviation)*

### Fit the model

```{r,warning=F}
fit_m2 <- sdmTMB(
  Sp10m3 ~ 0 + Year,
  data = data.eggs,
  mesh = mesh,
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "IID",   
  time = "time")

```

### Results of the fit

```{r,warning=F}
fit_m2
```

#### Confidence intervals of fixed effects

```{r,warning=F}
fix_effects<- tidy(fit_m2, conf.int = TRUE)
fix_effects
```

#### Confidence intervals of random effects and variance parameters

```{r,warning=F}
random_effects<- tidy(fit_m2, effects = "ran_pars", conf.int = TRUE)
random_effects
```

Other than the effect of the spatial random (sigma_O) and the spatiotemporal random field (sigma_E) field you have: :

-   Matérn range: the distance at witch you have no correlation - similar to the variogram range

-   phi (φ) & tweedie_p (p): the dispersion parameter and the power of the Tweedie distribution, respectively.

##### Plot the all the effects

```{r,warning=F}
ggplot() +
  geom_point(data = fix_effects, aes(x = estimate, y = term, col = term)) + 
  geom_errorbar(data = fix_effects, aes(y = term, xmin = conf.low, xmax = conf.high, color = term), width = 0.2) +  
  geom_point(data = random_effects %>% filter(term == c("sigma_O","sigma_E")), aes(x = estimate, y = term)) +  
  geom_errorbar(data = random_effects %>% filter(term == c("sigma_O","sigma_E")), aes(y = term, xmin = conf.low, xmax = conf.high), width = 0.2) +  
  labs( title = expression("Mean and 95% confidence intervals of the fixed & random effects," ~ m[2]),       
        subtitle = expression(mu[s~t] ~ "=" ~ exp(beta[t] + omega[s] +epsilon[st]))) +
  theme_bw() +
  theme(legend.position = "none")
```

You can see that the effect of spatial term is steal the most "important" one. But a lot of is effect is now explained by the spatiotemporal effect.

### Diagnostics

```{r,warning=F}

sanity(fit_m2)

```

### Prediction

You can steel predict in the same grid as m~1~ as no more covariates have been added to the model.

```{r,warning=F}
p_m2 <- predict(fit_m2, newdata = grid_yrs)
```

#### Full (fixed and random effects) model m~2~ prediction

```{r,warning=F}
plot_m2 <- ggplot(data=p_m2, aes(X, Y, fill = exp(est))) + 
  geom_raster() + facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (fixed effects +  random effects) based on full m2",
subtitle = expression("The estimated density of eggs " * log(mu[st]) * ": " * mu[st] == exp(beta[t] + omega[s]+ epsilon[st])))
plot_m2
```

#### Fixed effect in the prediction of model m~2~

```{r,warning=F}
plot_m2 <- ggplot(data=p_m2, aes(X, Y, fill = exp(est_non_rf))) + 
  geom_raster() + facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (fixed effects only) based on full m2",
          subtitle = expression("The effect of time in the estimated variable " * (beta[t]) * ": " * mu[st] == exp(beta[t] + omega[s]+ epsilon[st])))
plot_m2
```

#### Random effects in the prediction of model m~2~

Here spatial & spatiotemporal random effects

##### Plot the spatial random effect

```{r,warning=F}
plot_m2 <- ggplot(data=p_m2, aes(X, Y, fill = exp(omega_s))) + 
  geom_raster() +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (spatial random effect) based on full m2",
          subtitle = expression("The effect of space in the estimated variable " * (omega[s]) * ": " * mu[st] == exp(beta[t] + omega[s] + epsilon[st])))
plot_m2
```

##### Plot the spatotemporal random effect

```{r,warning=F}
plot_m2 <- ggplot(data=p_m2, aes(X, Y, fill = exp(epsilon_st))) + 
  geom_raster() +
  facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (spatiotemporal random effect) based on full m2",
          subtitle = expression("The effect of space through time in the estimated variable " * (epsilon[st]) * ": " * mu[st] == exp(beta[t] + omega[s] + epsilon[st])))
plot_m2
```

The spatiptempral random effect represent deviation from the fixed effect predictions and the spatial random effect deviations. These represent changes in the spatial paternal through time that are forced by parameters not accounted for in the model.

## MODEL 3 (m~3~) : Spatiotemporal model with covariates.

Now letts add a variable.

The 3^nd^ model (m~3~) will be a spatiotemporal model with Temperature as a covariate :

μ~s,t~= exp(X~s,t~β+β~t~+ ω~s~ + ε~s,t~),

μ~s,t~= log(y~s,t~ ),

y~s,t~ \~ Tweedie (μ~s,t~ ,p, φ) , ω\~ MVNormal (0,Σ~ω~) , ε~t~ \~ MVNormal (0,Σ~t~)

where the modeled egg density (μ~s,t~) follows the Tweedie distribution with a log link function, as egg density is positive, continuous and contains zeros.

y~s,t~ represent eggs density (no of eggs/10m^2^) at space (s) and time(t), μ is the mean density at time and space and p, φ represent power and dispersion parameters of the Tweedie distribution, respectively.

X~s,t,~ is the design matrix with covariates temperature and year as factor. Temperature is modeled with a smooth function.

β is a vector with the fixed effect coefficients (here only temperature)

β~t~ are the fixed effects for factor(Year), they represent the average temporal variation

ω~s~ is the spatial random effect, drawn from a GRF with covarianvce matrix Σ~ω~ using Maretn correlation function. It represents the average spatial pattern (nich effect) and it also accounts for spatial variation that is not accounted for by our fixed effects. In other words, these deviations represent consistent biotic and abiotic factors that are affecting biomass density but are not accounted for in the model.

ε~s,t~ represents the spatiotemporal random effect, drawn from a GRF with covarianvce matrix Σ~ε~. They represents how the average spatial pattern change through year so they also for spatial variation that is not accounted for by our fixed effects and the spatial random effects. In other words, these deviations represent consistent biotic and abiotic factors that are affecting our study variable but are not accounted for in the model.

In this model Spatiotemporal random fields are considered independent and identically distributed each single year (spatiotemporal = "IID") and Maretn correlation function to accounts for spatial autocorrelation.

To fit the new model we need to

-   Check the data are ok (mainly covariates)

-   Create new mesh based on the new data sets (some data points are removed)

-   Create a new prediction grid containing all the covariates (the [@lindmark2023] approach is used, see intro)

### Prepare the data

```{r,warning=F}
summary(data.eggs$Tempsbe)
hist(data.eggs$Tempsbe)
data.temp=data.eggs[data.eggs$Tempsbe>=0 & data.eggs$Tempsbe<30 & !is.na(data.eggs$Tempsbe),]
hist(data.temp$Tempsbe)
```

### Create the mesh

```{r,warning=F}
mesh <- make_mesh(data.temp, xy_cols = c("X", "Y"), cutoff = 30)
plot(mesh)
```

### Create the prediction grid

We first fit a spatiotemporal model (m~temp~) for temperature and then we predict temperature in the prediction grid that we already have. The results of this prediction will be the new prediction grid for the m~3~.

#### **MODEL Temp (m\~temp) : Spatio-temoral model for temperature**

The temperature model (m~temp~) follows that formula:

μ~s,t~= exp(β~t~+ ω~s~ + ε~s,t~),

μ~s,t~= y~s,t~ ,

y~s,t~ \~ Normal (μ~s,t~ σ^2^) , ω\~ MVNormal (0,Σ~ω~) , ε~t~ \~ MVNormal (0,Σ~t~)

where the modeled temperature (μ~s,t~) followe a Gaussian distribution without a link function.

y~s,t~ represents temperature at space (s) and time(t), μ is the mean density at time and space and and σ^2^ been the variance.

β~t~ are the fixed effects for factor(Year), they represent the average temporal variation

ω~s~ is the spatial random effect, drawn from a GRF with covarianvce matrix Σ~ω~ using Maretn correlation function.

ε~s,t~ represents the spatiotemporal random effect, drawn from a GRF with covarianvce matrix Σ~ε~. Spatiotemporal random fields are considered independent and identically distributed each single year (spatiotemporal = "IID") and Maretn correlation function to account for spatial autocorrelation.

##### Fit the model

```{r,warning=F}
fit_temp <- sdmTMB(
  Tempsbe ~ 0 + Year,
  data = data.temp,
  mesh = mesh,
  family = gaussian(link = "identity"),
  spatial = "on",
  spatiotemporal='IID',
  time='time'
  )

```

##### Results of the fit

```{r,warning=F}
fit_temp
```

##### Confidence intervals of fixed effects

```{r,warning=F}
tidy(fit_temp, conf.int = TRUE)
```

###### Confidence intervals of random effects and variance parameters

```{r,warning=F}
tidy(fit_temp, effects = "ran_pars", conf.int = TRUE)
```

###### Diagnostics

```{r,warning=F}
sanity(fit_temp)
```

##### Prediction

```{r,warning=F}
p_temp <- predict(fit_temp, newdata = grid_yrs)
```

```{r,warning=F}
plot_temp <- ggplot(data=p_temp, aes(X, Y, fill = est)) + 
  geom_raster() + facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "identity") +
    ggtitle("Prediction (fixed effects + random effects) based on m_temp",
subtitle = expression("The estimated temperatures " (mu[st]) * ": " * mu[st] == (beta[t] + omega[s]+ epsilon[st])))
plot_temp
```

This prediction output will be the prediction grid for m~3~.

### Fit the model m~3~

```{r,warning=F}
fit_m3 <- sdmTMB(
  Sp10m3 ~ 0 + Year + s(Tempsbe),
  data = data.temp,
  mesh = mesh,
  family = tweedie(link="log"),
  spatial = "on",
  spatiotemporal='IID',
  time='time'
  )
```

### Results of the fit

```{r,warning=F}
fit_m3
```

#### Confidence intervals of fixed effects

```{r,warning=F}
fix_effects<- tidy(fit_m3, conf.int = TRUE)
fix_effects
```

#### Confidence intervals of random effects and variance parameters

```{r,warning=F}
random_effects<- tidy(fit_m3, effects = "ran_pars", conf.int = TRUE)
random_effects
```

Other than the effect of the spatial random (sigma_O) and the spatiotemporal random field (sigma_E) field you have:

-   Matérn range: the distance at witch you have no correlation - similar to the variogram range

-   phi (φ) & tweedie_p (p): the dispersion parameter and the power of the Tweedie distribution, respectively.

##### Plot the all the effects

*The effect of Temperature is not in the fixed_effects because we used a smoother. How can I visualize this effect??*

```{r,warning=F}
ggplot() +
  geom_point(data = fix_effects, aes(x = estimate, y = term, col = term)) + 
  geom_errorbar(data = fix_effects, aes(y = term, xmin = conf.low, xmax = conf.high, color = term), width = 0.2) +  
  geom_point(data = random_effects %>% filter(term == c("sigma_O","sigma_E")), aes(x = estimate, y = term)) +  
  geom_errorbar(data = random_effects %>% filter(term == c("sigma_O","sigma_E")), aes(y = term, xmin = conf.low, xmax = conf.high), width = 0.2) +  
  labs( title = expression("Mean and 95% confidence intervals of the fixed & random effects," ~ m[3]),       
        subtitle = expression(mu[s~t] ~ "=" ~ exp(X[st]* beta + beta[t]+ omega[s] +epsilon[st]))) +
  theme_bw() +
  theme(legend.position = "none")
```

### Diagnostics

```{r,warning=F}
sanity(fit_m3)
```

### Prediction

```{r,warning=F}
p_temp = p_temp %>% rename(Tempsbe=est) 
p_m3 <- predict(fit_m3, newdata = p_temp)
```

#### Full (fixed and random effects) model m~3~ prediction

```{r,warning=F}
plot_m3 <- ggplot(data=p_m3, aes(X, Y, fill = exp(est))) + 
  geom_raster() + facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "log")+
    ggtitle("Prediction (fixed effects +  random effects) based on full m3",
subtitle = expression("The estimated density of eggs " * log(mu[st]) * ": " * mu[st] == exp(X[t,s]*beta +beta[t] + omega[s]+ epsilon[st])))
plot_m3
```

#### Fixed effects in the prediction of model m~3~

```{r,warning=F}
plot_m3 <- ggplot(data=p_m3, aes(X, Y, fill = exp(est_non_rf))) + 
  geom_raster() + facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (fixed effects only) based on full m3",
          subtitle = expression("The effect of time and temperature in the estimated variable " * (X*beta+ beta[t]) * ": " * mu[st] == exp(X[t,s]*beta +beta[t] + omega[s]+ epsilon[st])))
plot_m3
```

The pastern change is space - just because the effect of temperature is small and the scale is not good it is not reprisented well in the map.

#### Random effects in the prediction of model m~3~ - here spatial & spatiotemporal random effects

##### Plot the spatial random effect

```{r,warning=F}
plot_m3 <- ggplot(data=p_m3, aes(X, Y, fill = exp(omega_s))) + 
  geom_raster() +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (spatial random effect) based on full m3",
          subtitle = expression("The effect of space in the estimated variable " * (omega[s]) * ": " * mu[st] == exp(X[t,s]*beta+beta[t] + omega[s] + epsilon[st])))
plot_m3
```

##### Plot the spatotemporal random effect

```{r,warning=F}
plot_m3 <- ggplot(data=p_m3, aes(X, Y, fill = exp(epsilon_st))) + 
  geom_raster() +
  facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "log")+ 
  ggtitle("Prediction (spatiotemporal random effect) based on full m3",
          subtitle = expression("The effect of space through time in the estimated variable " * (epsilon[st]) * ": " * mu[st] == exp(X[t,s]*beta+beta[t] +omega[s] + epsilon[st])))
plot_m3
```

#### Visualize the effect of the smoothed covariate Temperature.

How I do not get that ???

```{r,warning=F}

nd <- data.frame(
  Tempsbe = seq(min(data.temp$Tempsbe), max(data.temp$Tempsbe), length.out = 100
  ),
  Year = as.factor(2024),time=24 # a chosen year
)

p <- predict(fit_m3, newdata = nd, se_fit = TRUE, re_form = NA)

ggplot(p, aes(Tempsbe, exp(est),
  ymin = exp(est - 1.96 * est_se),
  ymax = exp(est + 1.96 * est_se)
)) +
  geom_line() +
  geom_ribbon(alpha = 0.4) +
  scale_x_continuous() +
  coord_cartesian(expand = F) +
  labs(x = "Temp (°C)", y = "Density (Nb/10m3)")

```

# Prepare more covariates for more models ....

## MODEL prof (Prof) : Model spatio-temporel de Covariables

### Prepare covariable

```{r,warning=F}
data.prof=data.eggs[data.eggs$Sonde>=0 & data.eggs$Sonde<5000 & !is.na(data.eggs$Sonde),]
hist((data.prof$Sonde))
```

### Create the mesh

```{r,warning=F}

mesh <- make_mesh(data.prof, xy_cols = c("X", "Y"), cutoff = 30)
plot(mesh)

```

### Fit the model

```{r,warning=F}

fit_prof <- sdmTMB(
  Sonde ~ 0 + Year,
  data = data.prof,
  mesh = mesh,
  family = gaussian(link = "log"),
  spatial = "on",
  spatiotemporal='IID',
  time='time'
  )

```

### Results of the fit

```{r,warning=F}
fit_temp
```

```{r,warning=F}

tidy(fit_temp, conf.int = TRUE)

```

```{r,warning=F}

tidy(fit_temp, effects = "ran_pars", conf.int = TRUE)

```

### Diagnostics

```{r,warning=F}

sanity(fit_temp)

```

### Prediction

```{r,warning=F}

p_prof <- predict(fit_prof, newdata = grid_yrs)

```

```{r,warning=F}

plot_prof <- ggplot(data=p_prof, aes(X, Y, fill = exp(est))) + 
  geom_raster() + facet_wrap(vars(Year)) +
  scale_fill_viridis_c(trans = "log")
plot_prof

```
